---
title: "R Notebook"
output: html_notebook
---
Analyzing growth. First setup the dataframe that will be used. We load the libraries we will use and read in the csv file. 

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
library(reshape2)

growth <- read.csv(file='C:/Users/Ruben.Sikkes/Downloads/Growth.csv', header=T)

# up until colomn 12 because after is for now not needed and exclude empty rows at the bottom
growth <- growth[3:(which(is.na(growth$Weight))[1]-1),1:12]

# parse date and create a column for days.
growth$Date <- parse_date_time(growth$Date, "%d/%m/%y")
growth$Wday <- wday(growth$Date)
growth$day_since_start <- as.double(difftime(growth$Date, growth$Date[1], units = 'days'))
head(growth)
```
Now for a table view here's an indication of raw and percentual changes measured. How has my composition and raw stats changed?

```{r}
# composition beginning vs end
change_df_pct = data.frame("fat_pct" = c(growth$fatpct[1], growth$fatpct[length(growth$day_since_start)]),
                      "muscle_pct"= c(growth$musclepct[1], growth$musclepct[length(growth$day_since_start)]),
                      "wat_pct"= c(growth$watpct[1], growth$watpct[length(growth$day_since_start)]))
change_df_pct[3,] <- change_df_pct[2,]-change_df_pct[1,]
row.names(change_df_pct) <- c("Before", "Current", "Change")
head(change_df_pct)
```

```{r}
change_df_raw <- data.frame("fatmass" = c(growth$fatmass[1], growth$fatmass[length(growth$day_since_start)]),
                      "musclemass"= c(growth$musclemass[1], growth$musclemass[length(growth$day_since_start)]),
                      "watmass"= c(growth$watmass[1], growth$watmass[length(growth$day_since_start)]))
change_df_raw[3,] <- change_df_raw[2,]-change_df_raw[1,]
row.names(change_df_raw) <- c("Before", "Current", "Change")
head(change_df_raw)
```

Now to plot some basic things as how weight changes over time.

```{r}
ggplot(growth, aes(x=day_since_start,y=Weight)) + geom_point() + geom_smooth()+ theme_minimal()+
  labs(x="Day since start", y = "Weight (Kg's)", title="Weight over time", 
       subtitle= paste("Weigth gained is ", 
                       round(growth$Weight[length(growth$Weight)]-growth$Weight[1], digits=2),
                       "KG over ",max(growth$day_since_start),"days"))
```
We can see that weight increases and is starting to stagnate.

```{r}
ggplot(growth, aes(x=day_since_start,y=musclemass)) + 
  geom_point() +
  geom_smooth(color='green')+ theme_minimal()+
  labs(x="Days since start", y = "Musclemass (Kg's)", title="Musclemass over time",
  subtitle= paste("Muscle gained is ", round(growth$musclemass[length(growth$musclemass)]-growth$musclemass[1], digits=2),
                       "KG over ",max(growth$day_since_start),"days"))
```
Similarly we can see that musclemass increases but stagnates. Idem for fatmass below.

```{r}
ggplot(growth, aes(x=day_since_start,y=fatmass)) + geom_point() + geom_smooth(color='red')+ theme_minimal()+
  labs(x="Days since start", y = "Fatt (Kg's)", title="Fatmass over time", subtitle= paste("Fat gained is ", round(growth$fatmass[length(growth$fatmass)]-growth$fatmass[1], digits=2),
                       "KG over ",max(growth$day_since_start),"days"))
```
Now taking into consideration the gain per day of the different variables, we create new variables showing the difference compared to timestep 0 and create a new melted dataframe for plotting. 
```{r}
growth$total_gain <- growth$Weight- growth$Weight[2]
growth$musc_total_gain <- growth$musclemass- growth$musclemass[2]
growth$fat_total_gain <- growth$fatmass- growth$fatmass[2]
growth$wat_total_gain <- growth$watmass- growth$watmass[2]

# makes it easier to plot (with correct legend), we take the newly created last 4 variables
temp <- growth[2:nrow(growth),15:18]
temp2 <- melt(temp)
temp2$day <- rep(growth$day_since_start[2:nrow(growth)], times=4)

ggplot(temp2) + 
  geom_smooth(aes(x=day, y= value, colour=variable), se=F, method = "loess") +
  labs(title= "Gain over Time", x= "Day", y = "Gain this day (kg)", 
       subtitle= paste( "Max weight difference since start is " , round(max(growth$total_gain[2:nrow(growth)]),3), "Kg")) +
  geom_hline(aes(yintercept = 0, linetype="no gain")) +
  scale_linetype_manual(name = "Reference",values = c(2, 2)) +
  theme_minimal()
```

Now looking more specifically at weight change per day we get the following graph. 
```{r}
day_length= length(growth$Date)
dd_sub = growth[2:day_length,c(8,10,12)]
day_length= length(growth$Date)
dd = melt(dd_sub)

dd$day <- rep(growth$day[2:day_length], times=3)

ggplot(dd) + 
  geom_smooth(aes(x=day, y= value, colour=variable), se=F) +
  labs(title= "Gain per day", x= "Day", y = "Gain this day (kg)") +
  geom_hline(aes(yintercept = 0, linetype="no gain")) +
  scale_linetype_manual(name = "Reference",values = c(2, 2)) +
  theme_minimal()
```
We quickly notice that smoothed gain has been good untill around day 65. After day 65 i seem to be showing losses. Good though is that most extreme losses are water and not muscle!

Side question: is there a difference in day of the week and gain?
```{r}
growth%>% group_by(Wday) %>% summarize(sum= sum(muscgain)) 

```
```{r}
growth%>% group_by(Wday) %>% summarize(summed_gain= sum(muscgain)) %>% ggplot() + 
  geom_col(aes(x=Wday, y= summed_gain, fill=summed_gain))+
  scale_fill_gradient(low ="red", high= "green")+ theme_minimal() + 
  labs(x= "Day of the Week", y = "Summed Gain", title="Muscle Gain per Day of the Week")+
  xlim(labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
```
So sunday is the day that really fucks me over. Which makes sense because it is typically the day after I go drinking. Wednesday on the other hand is normally the day after which I have not gone to the gym for 2 days. So maybe this hints that muscle growth only works 2 days after a workout?


Next: how will my weight change the next 14 days? We use a simple regression model based on weight and time. 

```{r}
# create the model which we will use 
fit <- lm(Weight~poly(day_since_start,1)+ poly(day_since_start,2),data=growth)

#summary(fit) # display model results if you want

# Create a dataframe for prediction 
new_x <- as.data.frame(day_since_start<- max((growth$day_since_start)+1):(max(growth$day_since_start)+14))

# predict new values
predicted_weight <- predict(fit, new_x)

# and plot results
ggplot() + geom_point(aes(x=new_x, y = predicted_weight))+ geom_smooth(aes(x=new_x, y = predicted_weight))+
  labs(x='Day since start', y='weight', title = 'Predicted Weight for Next 14 Days') + theme_minimal()

#tail(predicted_weight) #  show predicted results if you like
```

```{r}
ggplot() + geom_point(aes(x=day_since_start, y = Weight), data=growth) +
  geom_smooth(aes(x=day_since_start, y = Weight),color='blue', data=growth, method ='loess') +
  geom_smooth(aes(x=new_x, y = predicted_weight), color='purple')+
  theme_minimal()+
  labs(title='Measured and Predicted compared to 85kg goal', x="Day since start", y = "Weight (kg's)")+
  geom_hline(yintercept=85, color='red')+
  scale_x_continuous(breaks=seq(0,100,10))
```
Based on this figure we can see that I am never reaching 85 based on my current trajectory. Current trajectory indicates my weight will decrease in a couple of days.

So how about the ratio between fatgain and muscle gain? How has my bulk been? We can visualize this by the ratio of Musclegain/Fatgain.
```{r}
growth$mf_gain_ratio <- growth$muscgain/ growth$fatgain
growth %>% filter(muscgain>0) %>% ggplot(aes(x=day_since_start, y= mf_gain_ratio)) +
  geom_point() +
  geom_smooth(se=F)+
  theme_minimal()+
  scale_y_continuous(breaks = seq(0,3,0.5) )+
  labs(x="Day since start", y= "muscle to fatgain ratio", title='Muscle to Fatgain ratio over time', subtitle= "based on good= m:f > 1:1, bad worse than 1:2 and average in between")+
  geom_rect(xmin=0, ymin=0, xmax=max(growth$day_since_start),ymax=0.5, colour="red", fill="red", alpha=0.005)+
  geom_rect(xmin=0, ymin=1, xmax=max(growth$day_since_start),ymax=3, colour="green", fill="green", alpha=0.005)

```
Glancing at the results we see that at as hoped, most bulking days score atleast above a 1:1 ratio (above the red box)

Next we show that comparing how many dots are in each box we see that most is in blue (middle) and thereafter green (good).
```{r}
bad_pct = sum(growth$mf_gain_ratio<0.5, na.rm=T)/ sum(growth$mf_gain_ratio>0, na.rm=T)
good_pct = sum(growth$mf_gain_ratio>1, na.rm=T)/ sum(growth$mf_gain_ratio>0, na.rm=T)
middle_pct= sum(growth$mf_gain_ratio>0.5 & growth$mf_gain_ratio<1, na.rm=T)/ sum(growth$mf_gain_ratio>0, na.rm=T)

gain_df = data.frame('ratio'= c(bad_pct,middle_pct,good_pct), group=c('bad','middle', 'good'))

ggplot(data=gain_df, aes(x=group, y=ratio, fill=group)) + geom_col() +
  labs(x= "", y='Overall part', title='Muscle to Fat Ratio Group Percentages', subtitle= paste("mean ratio = ", mean(gain_df$ratio))) + theme_minimal()
```
Overall my bulk has therefore been quite oke since overall my mean ratio is around 0.36. Which is visualized also in the density plot below.
For completeness we show a histogram of how often each value is reached. We are on the right side of the vertical line atleast.
```{r}
ggplot(data= dd, aes(x=value, fill=variable)) + 
  geom_density(alpha=0.2) + 
  geom_vline(xintercept = 0)+
  theme_minimal() + 
  labs(x='Gain', title= 'Density for Gain compared to nothing (x=0)')
```

So whats the relation between eating more and gain? Does having a bigger surplus indeed result in a worse muscle/fat gain ratio? 
```{r}
growth %>% filter(muscgain>0) %>% ggplot() + geom_point(aes(x=fatgain, y = muscgain, colour= mf_gain_ratio)) + 
  geom_smooth(aes(x=fatgain, y = muscgain), colour="black", method ="lm", se=F) + 
  geom_abline(slope=1, colour= "green", linetype=2) + 
  geom_abline(slope=0.5, colour="red", linetype=2)+
  labs(title= "The more you gain the worse the muscle/fat ratio becomes", subtitle="green= 1:1, black= average measured, red= 1:2") + theme_minimal() + scale_colour_gradient(low ="red", high= "green")
```
The best points are definately in the left-bottom corner and moving to the right, the black line moves closer to the red dashed line (a bad ratio). Gain should, for an ideal ratio, therefore be minimized. 


Lastly an overall indication of physical fitness is the FFMI ratio. For this we use the following formulas:
Lean Weight = Weight * (1 - (body fat % / 100)
FFMI = (Lean Weight / 2.2)/ ((Feet * 12.0 + Inches) * 0.0254)2
Adjusted FFMI = FFMI + ( 6.3 * (1.8 - (Feet * 12.0 + Inches) * 0.0254))

```{r}
growth$lean_weight <- growth$Weight * (1-(growth$fatpct /100))
growth$ffmi <- (growth$lean_weight/2.2)/(((6*12)+4)*0.0254)^2*2.20462
growth$adj_ffmi <- growth$ffmi + (6.1 * (1.8 - ((6 * 12.0 + 4) * 0.0254)))

ggplot(growth, aes(x=day_since_start,y=ffmi)) + geom_point() + geom_smooth()+ theme_minimal()+
  labs(x="Day since start", y = "ffmi", title="ffmi over time", 
       subtitle= paste("FFMI Improved is", 
                       round(growth$ffmi[length(growth$ffmi)]-growth$ffmi[1], digits=2),
                       "FFMI over ",max(growth$day_since_start),"days"))

ggplot(growth, aes(x=day_since_start,y=adj_ffmi)) + geom_point() + geom_smooth()+ theme_minimal()+
  labs(x="Day since start", y = "adjusted ffmi)", title="Adjusted FFMI over time", 
       subtitle= paste("adj_ffmi gained is ", 
                       round(growth$adj_ffmi[length(growth$adj_ffmi)]-growth$adj_ffmi[1], digits=2),
                       "adj_ffmi over ",max(growth$day_since_start),"days"))
```

There fore based on FFMi indexes:  
* 16 - 17: below average   
* 18 - 19: average  
* 20 - 21: above average  
* 22 - 23: excellent  
* 23 - 25: superior  
* 26 - 27: scores considered suspicious but still attainable naturally  
* 28 - 30: highly unlikely scores to be obtained naturally without steroid usage  

still average since im around 18.2 for normal and even below average for adj_ffmi. 
